Intracellular mechanical fingerprint reveals cell type specific mechanical tuning

Version 1

Abstract

Abstract

Living cells are complex entities that perform many different complex tasks with astonishing robustness. While the direct dependence of biological processes on controlled protein expression is well established, we only begin to understand how intracellular mechanical characteristics guide and support biological function. This is in stark contrast to the expected functional role that intracellular mechanical properties should have for many core cellular functions such as organization, homeostasis and transport. From a mechanical point of view, cells are complex viscoelastic materials that are continuously driven out of thermodynamic equilibrium, which makes both a physical measurement and mathematical modeling of its properties difficult. Here, we define a “mechanical fingerprint” that can not only characterize the intracellular mechanical state, but also carve out the mechanical differences between cell types with the potential to relate these to proper cell function. By analyzing the frequency-dependent viscoelastic properties and intracellular activity of cells using microrheology, we distilled the complex active mechanical state into just 6 parameters that comprise the mechanical fingerprint. The systematic investigation of the fingerprint illustrates a parameter tuning that can be explained by the functional cellular requirements. However, the full potential of the mechanical fingerprint is given by a statistical analysis of its parameters across all investigated cell types, which suggests that cells adjust mechanical parameters in a correlated way to position their intracellular mechanical properties within a well defined phase-space that is spanned between activity, mechanical resistance and fluidity. This paves the way for a systematic study of the interdependence of biological function and intracellular active mechanics.

Introduction

Cells are highly complex, heterogeneous materials that tightly control their viscoelastic characteristics by precise tuning of the active mechanical properties to maintain a non-equilibrium mechanical steady state. This mechanical homeostasis requires a continuous consumption of metabolic energy, mostly in the form of ATP and is maintained while the system undergoes various dynamic functions ranging from tissue morphogenesis and patterning1, 2 to organelle positioning,3 differentiation,4 migration5, 6 and polarization.7

Although it is well-acknowledged that intracellular mechanical properties are key to understanding major cellular functions, most studies have measured cellular mechanics from outside of the cell using e.g. AFM,8, 9 micropipette aspiration,10 traction force microscopy,11 tether pulling,12 magnetic twisting cytometry13 or deformability cytometry.14 While these studies allowed deep insights into the relevance of cell mechanics for supracellular biological processes by probing predominantly the stiff actin cortex that surrounds the cell, inferring information about the cytoplasm mechanics remained difficult.15 This experimental gap has recently been bridged by active intracellular rheology experiments that are based on optical as well as magnetic tweezers.16–19 For example, a recent study has shown that the cytoplasm is softening during mitosis20 which might be key for the proper functioning of this process.

While a large body of work has focused on the viscoelastic properties of cells, it became increasingly clear that the well-controlled active force generation of living cells has to be integrated to obtain a full intracellular mechanical description. A multitude of recent studies has therefore focused on investigating these non-equilibrium, active properties of cells.21–23 These active, energy consuming cellular processes are a key component to maintain the intracellular organization despite the ever-dispersing forces of entropy.16, 24 While the intracellular transport continuously organizes membrane and organelle architecture via motor activity, the dynamic growth and disassembly of cytoskeletal filaments ensures global processes such as migration,25 proliferation26 or wound healing.27 Further, recent studies have shown that dynamic waves of actin filaments also organize mitochondria position during cell division28 and that cytoplasmic forces can reorganize nuclear condensates,29 thus emphasizing the key importance of active forces for fundamental cellular function.

Despite the generally accepted relevance of intracellular active mechanics for the functioning of entire organisms, it remains difficult to quantify and even harder to model theoretically. From a physical perspective, cells behave as viscoelastic materials where timescales are a key information to correctly identify the dominating mechanical principles. Therefore, a full quantitative description requires microrheological information, typically provided in the form of a frequency dependent, complex shear modulus G∗(f).30 Although this quantification of the viscoelasticity is experimentally accessible, its interpretation requires theoretical models to become physically meaningful and to reduce the complexity of data. However, while simple mechanical anologies such as Kelvin-Voigt or Maxwell models lack the complexity to describe the observed behaviour, more elaborate models such as chains of springs and dashpots require many parameters and thus do not reduce the complexity significantly. This has led to double power law approaches, that are motivated by fractional Generalized Kelvin-Voigt (fGKV) models.31 In recent work these have been successfully used to quantify the viscoelastic properties of fibroblasts,32 endothelial33 and epithelial cells20 and they reduce the complex mechanical behaviour to just four parameters. Recent efforts to include the active force generation in a mechanical characterization have been able to describe the active energy injected into the cellular system also by a power law20 that is defined by two additional parameters.

These insights beg the question to which extent living cells actively control their intracellular mechanical properties, or if these are simply the byproduct of the cytoplasmic composition. Assuming a functional relevance, the parameters may vary considerably between cells, and might even allow to identify different cell types and cellular situations in a fingerprint-like fashion. The relevance of this 6 parameter description was recently established by showing systematic changes during cell division, and to perform new statistical analysis of stochastic particle trajectories.20, 34 In this study we use optical tweezers based active microrheology and passive observation of intracellular particle motion to demonstrate that cell types can be successfully distinguished in a pairwise comparison to position cells on 3D phasespace that reflects activity, rigidity and fluidity.

Mechanical parameter set quantifies active mechanics

Intracellular viscoelastic material properties are commonly quantified by a frequency-dependent, complex shear modulus that we directly determine via optical tweezers based active microrheology (M 2). Here, a custom-built optical tweezers setup (Figure 1a, M 1) exerts well-defined oscillatory forces onto phagocytosed 1 µm sized probe particles inside cervical cancer cells (HeLa). To determine the resulting bead displacement, a stationary, weak detection laser monitors the particle position with nanometer precision. Applying sinusoidal forces (Figure 1b) with frequencies between 1 Hz and 1024 Hz allows to calculate the mechanical response function χ(f), which is used to determine the complex shear modulus G∗(f) = 1/(6πRχ(f)), with R the probe particle radius (Figure 1c). G∗(f) is a complex quantity where the real part G′(f) (storage modulus) describes the elastic properties of the cell and the imaginary part G′′(f) (loss modulus) describes the viscous properties (Figure 1c).

a) Viscoelastic properties and activity are measured within the cytoplasm of cells using a custom build optical tweezers setup. b) Sinusoidal forces (blue) at varying frequencies are applied to a probe particle while the particle displacement (red) in response to this force is monitored. c) Repeating this procedure at different frequencies allows to calculate viscoelastic material properties in terms of the complex shear modulus G∗(f) which consists of the storage modulus (blue) and a loss modulus (red). Both storage and loss modulus can collectively be fitted using a generalized fractional Kelvin-Voigt (fGKV) model(inset). d) This fGKV model consists of two power laws which in a simplified way can be pictured as a more solid-like material class resembling polymeric filaments and a more fluid-like material class (crowded molecules). e) Using an additional passive measurement, violation of the fluctuation-dissipation theorem can be directly visualized. (red area) f) The effective energy quantifies intracellular activity and can be fitted with a two parameter power law. g) Complex and frequency dependent intracellular active mechanical properties can be reduced to a fingerprint of just 6 parameters, which describe the intracellular mechanical state.

To model the shear modulus we follow the naive hypothesis that the cytoplasm consists of two main material classes, namely an elastic solid-like material that is generated by a sparse, connected polymer network, and a more liquid-like material that fills the bulk space of the cytoplasm (Figure 1d). Generally, such a combination of two complex materials can be approached by a linear combination of two springpots that leads to a double power law model consistent with classical rheology data20, 31–33 : This complex function can then be used to simultaneously fit both the real and imaginary part of the experimentally obtained shear modulus to determine the prefactors and power law exponents. As shown in Figure 1c the model fits the experimental data well. R2-values of the fits are given for raw trajectories (Figure 2, Table 14) and bootstrapped trajectories (Figure 3, Table 16). Consistent with the initial hypothesis, we identify a region with an power law exponent between 0 and 0.5 that we attribute to a more solid-like material (ideal solid has exponent 0), and a second power law that is between 0.5 and 1 (ideal fluid has exponent 1) corresponding to a liquid-like material.

Wild type HeLa cells (WT) were treated with different cytoskeletal drugs to investigate their influence on the fingerprint parameters. a) Cytochalasin B (CytoB/CB) inhibits actin polymerization, Nocodazole (Nocoda/Noc) disturbs microtubule assembly and the combined treatment was used to disturb actin and microtubules simultaneously. b) The treatment with CB does not show a strong effect on the fingerprint parameters. c) Using Noc, intracellular activity in terms of E is strongly decreased while also the parameter B increases. d) The combined treatment shows a strong effect in decreasing intracellular activity E but also drastically softens the cytoplasm which is reflected in a decreased factor A. e,f,g) Change of fingerprint parameters in response to drug treatment. h,i,j) Effect of pharmacological treatment shown for the elastic modulus, loss modulus and effective energy

a) The mechanical fingerprint of 7 different cell types was determined by measuring their complex shear modulus and effective energy. b) Comparing HeLa cells to C2C12 muscle cells shows that muscle cells are overall stiffer (higher A). c) The comparison between HeLa cells and macrophages shows that macrophages of a higher intracellular activity (increased E) but are also more liquid-like and softer (increased α, β and decreased B) d-f). Using the fingerprint, active mechanical difference between cell types can be compared according to changes in the more solid-like material properties, more liquid-like material properties and the intracellular activity.

Besides intracellular mechanics, the generated active forces are another key determinant of the intracellular state. As cells consume metabolic energy to drive motor proteins and many other cell biological functions, cells are far from thermodynamic equilibrium. Experimentally, this is directly reflected in the motion of tracer particles within the cell. In thermodynamic equilibrium, the random motion of particles depends only on material properties and temperature, which is manifested by the fluctuation-dissipation theorem:35 Here, C(f) is the power spectral density, quantifying the spontaneous fluctuations of the particle, χ(f) is the mechanical response function and kBT is the thermal energy. In active, living cells particle mobility is beyond motion explainable by thermal fluctuations36 (Figure 1e), as cytoskeletal rearrangement and transport processes also contribute to the motion of probe particles. An elegant way to summarize all active and thermal forces is by introducing an effective energy EEff 35(M 2). Intuitively, this effective energy reflects the ratio between observed particle motion C(f) thermodynamically predicted motion that reflect the dissipative material properties as quantified by χ′′(f): It should be noted that this idea of an effective energy implies that the active forces generated by the cell share statistical characteristics with thermal forces. In this sense we use the effective energy rather as a phenomenological quantity that measures the extent of non-equilibrium mechanical driving of the system. The free particle fluctuation spectrum C(f) is acquired by turning off the trapping laser while using the lower-power detection laser to monitor the particle position. The resulting average effective energy measured for HeLa cells (Figure 1f) shows an increase influence of active forces in the low frequency regime, while the effective energy converges to the thermal limit of 1 kBT for higher frequencies. This is a direct demonstration that metabolic energy dominates slow particle motion, while assuming a thermodynamic equilibrium from shorter timescales is an adequate approximation. This asymptotic behaviour at short time scales is to be expected as motor proteins operate at longer time scales,37 and has previously been shown experimentally.16 Interestingly, this suggests that the question about non-equilibrium processes can be reduced to a question of timescales, which is a potential explanation why this subject is so controversially discussed in literature. This asymptotic behaviour offers an additional calibration method of optical tweezers in complex living systems35 that we used to optimize the readout of the effective energy and mechanical properties (see SI 1.1 for further discussion).

Similar to the mechanical properties, the effective energy can also be fit by a power law approach (Figure 1f): This enabled us to use only two parameters when quantifying all active forces injected in the cell by metabolic processes in a single description. In combination with the 4 parameters that are sufficient to describe the viscoelastic properties, we can establish a set of 6 parameters that fully describes the state of a cell, and in particular the active mechanics of the cytoplasm in HeLa cells (Figure 1f). As we show in the following, the set of these 6 parameters can be interpreted as mechanical fingerprint of a cell (Figure 1g).

In this study, we explore the variation of the mean active mechanical quantities G′, G′′, and EEff across different cell types or conditions. To obtain a robust and reliable distribution of these curves, we apply a bootstrapping scheme to the recorded single cell data (see M 3.1). The bootstrapped data is then subjected to respective model fitting, providing estimations for the distributions of the mean fingerprint parameters for each condition. Unless otherwise specified, all presented data is based on the bootstrapped data.

The cytoskeleton selectively affects the mechanical fingerprint

After verifying that HeLa cells are well described by the mechanical fingerprint, we wondered if different parameters can be affected by pharmacological perturbation of the cytoskeleton (Figure 2a). Commonly, the mechanical properties of cells are attributed to the actin cytoskeleton, which forms the cell cortex and other structures. Consequently, we disrupted the actin cytoskeleton in Hela cells using Cytochalasin B (CytoB or CB) and investigated the resulting fingerprint. To our surprise, we did not observe strong changes in intracellular active mechanics (Figure 2b-j grey). We solely observe a weak, yet significant decrease of solid-like exponent α (0.36 to 0.31). Neither the elastic modulus nor the viscous modulus of the cytoplasm (Figure 2h,i grey) is visibly affected by disruption of the actin cortex. As the probe particle is phagocytosed, it is rapidly transported away from the cell surface, and typically confined in a late endosome or lysosome. This can explain why the viscoelastic properties in terms of fingerprint parameters A, α, B and β experienced by the probe particle are mostly unaffected and why the effective energy remains unaffected (Figure 2 b-j grey). As the interaction of phagocytosed particles is mainly dominated by microtubule associated motor proteins, no effects of actin perturbation are expected. This explanation predicts that both, the effective energy and the viscoelastic material properties, should be affected when depolymerizing microtubules (MT). To test this prediction, we disrupted the MT network using 10 µg mL−1 Nocodazole (Nocoda or Noc). As expected, the activity was almost completely abolished (Figure 2 c,g,j yellow) by this treatment as mainly captured by a drastic reduction of the effective energy defining prefactor E. Strikingly, even depolymerization of MT did not reduce the contribution of the solid-like material properties A significantly. This result suggests that a yet unknown control mechanism stabilizes the intracellular mechanical properties of the solid-like material contribution. Cross-talk and feedback-loops between cytoskeletal structures are well known.38 A possible mechanism could be the known up-regulation of actin polymerization upon MT depolymerization.39, 40 If the stability of the prefactor A, the solid-like contribution, is due to a compensation mechanism where actin compensates MT, a double disruption should have a drastic effect, as any in-built compensation is suppressed. Indeed, when disrupting the actin and the MT network simultaneously, we do find the expected drop in both the active forces driving particle fluctuations, and the contribution of the solid-like material marked by a collapse of the prefactor A (Figure 2d,e,g blue).

Additionally, both treatments with Nocodazole are also accompanied by a significant increase in liquid-like prefactor B which indicates a yet not understood change in the cytoplasmic structure. Since the prefactor A is in general almost a magnitude larger than B, the overall viscoelastic properties are mainly dominated by A.

These results demonstrate not only that the mechanical fingerprint can be used to dissect the effect of cytoskeletal drugs on different properties but also the combined information obtained is a first hint that cells may use compensation mechanisms to regulate their intracellular mechanics.

Separating cell types by their mechanical fingerprint

As we have shown that these drastic pharmacological treatments largely affect the prefactor A of the solid-like material and the strength of the effective energy E driving intracellular fluctuations, we asked if these are indeed tuned by the cell to support certain biological functions. To test this idea, we decided to compare HeLa cells to two other cell types that have drastically different mechanical requirements according to their function. First, we hypothesized that a muscle cell needs to have an increased intracellular mechanical stiffness for its function as mechanical force generator. This suggests that muscle cells will be tuned to differ in the solid-like material properties with higher prefactor A.

As seen in Figure 3b, this is exactly what we found when comparing the cervical cancer HeLa cells with a murine derived myoblast cell line (C2C12). The solid-like material stiffness as quantified by A almost doubled, indicating a strong increase in intracellular resistance. Additionally the solid-like material exponent further decreased emphasising a more solid-like material (Figure 3b,d). All other fingerprint parameters did not change significantly.

Next we hypothesized that a fast-moving cell type would have increased active forces in the cytoplasm, and render it more liquid-like to be able to move quicker through the extra-cellular space. This hypothesis was tested using an immune cell model. Macrophages were obtained from a hematopoietic progenitor cell line that expresses HoxB8 under an estrogen promotor.11,13 Upon removal of estrogen the cells differentiate within 3 days to macrophages that we term HoxB8 cells in the following for convenience. Again, our experimental analysis of the mechanical fingerprint perfectly confirmed the prediction based on the biological function of macrophages (Figure 3c). Firstly, the effective energy prefactor E more than doubled indicating an increased intracellular activity. Secondly, there are changes in most parameters describing the mechanical properties. Both exponents α and β are significantly increased. Strikingly, the viscosity defining prefactor B of the liquid-like material property collapsed indicating that the viscoelastic properties can quickly dissipate any large scale deformation of the whole cell. This suggests that both solid- and fluid-like material components are tuned towards more liquid-like materials with a lower viscous component. Our result is in perfect agreement with measurements obtained previously on the whole cell level.41

These results furthermore support the hypothesis that cells are able to adjust their mechanical fingerprint depending on their biological function. Muscle cells, requiring higher mechanical resistance, show a large solid-like prefactor, while immune cells that need to migrate quickly and deform easily are more liquid-like with lower viscous dissipation and generate a larger cytoplasmic active energy.

The mechanical fingerprint is different across cell types and species

Motivated by the finding that different cell types have different mechanical properties as quantified by the mechanical fingerprint, we wondered whether this approach can be used as a strategy to differentiate between cell types in terms of mechanical properties. In detail, we ask if at least one of the 6 parameters of the mechanical fingerprint are statistically different when comparing different cell types. Hence, we decided to determine the 6 parameter fingerprint for 7 different cell types (Figure 3a), of various function and from various species. In addition to the already introduced human HeLa, murine C2C12 and HoxB8 cells, we measured the fingerprint of human lung epithelial cells (A549), a murine fibroblast-like colon carcinoma cell line (CT26), a canine kidney epithelial cell line (MDCK) and a non-invasive human breast cancer cell line (MCF7). Figure 3d-f shows the variations of the 6 fingerprint parameters across all seven cell types.

In a first analysis, we looked at all possible pair-wise combinations of cells to test our hypothesis that the 6 parameters are sufficient to differentiate between cell types. Indeed, in most of the 21 possible combinations we found at least one of the fingerprint parameters to be different (Figure 4a, SI 2.2). Interestingly, the combinations of cell types where no significantly different parameter could be found are epithelial cells MDCK, HeLa and MCF7 as well as the fibroblast CT26. On average 2.1 parameters showed a significant difference. However, this analysis does not take into account the extent of significant difference. To approach this question more systematically, we use a variant of the z-score, where we normalize the difference in the mean values by twice the geometric average of the standard deviation the bootstrapped data. The z-score hence quantifies the extend of deviation between cells that is carried by a single parameter. For example the z-score of A between HeLa and CT26 is:

a) Number of significant different fingerprint parameter in pairwise cell type comparisons. b) Frequency of parameters showing the highest z-score. c) Distribution of z-score for all parameters for pairwise cell comparison. d) Correlation analysis between fingerprint parameters shows that not all parameters are varied independently. e) left: Explained variance ratio of the different principal components. Almost 80% of the variance is explained using the first two components. right: Relative contribution of the fingerprint parameter to the first two principal components. C1 mainly consists of α, β and B, C2 mainly consists of A, E, γ. f) Plotting all cell types according tothe first two principal components C1 and C2 shows that two parameters are already sufficient to distinguish between most cell types. g) Qualitative phase diagram of the active mechanical space. Activity, mainly captured by parameter E0, resistance, dominated by A and solid-liquid switching can be described as fluidity that is determined by principal component 1 (C1) are varied among different cell types. Using this three dimensional space allows to identify physical differences among different cell types which may be related to function.

Here, any value larger than 1 can be considered to mark a significant difference, and the larger the z-score the better the two cells can be separated according to the respective parameter. We then calculated the z-score for all 6 fingerprint parameters across all 21 different pairs of cells. It should be noted that, in contrast to the p-value of the t-test, this values depends only on the first moments of the probability distributions, and is independent of the number of samples. Hence it reflects the average properties of each cell type, rather than a statistical measure. In a first step we studied which of the parameters was best to differentiate between the cells. Figure 4b shows how often each parameter was the best discriminator. We see that E and A are the most important parameters to distinguish between different cell types as they together show the highest z-score in more than almost 80% of the cases. As they mainly characterize the solid-like material properties and overall activity, this result implies that these properties are highly dependent on cell type and function. Additionally, we directly plot the z-scores of all cell pairs for the different parameters (Figure 4c). In contrast to the previous quantity, here the value of the z-scores becomes important. Despite E and A having most frequently the highest z-score, we also observed that the average z-score for the parameters β, α, A, γ and B was similar or smaller than 1, suggesting that the combination of all 6 parameters is relevant for characterizing the mechanical state of cells. The similarity in average z-scores may also indicate that there are correlations between these parameters, which could be due to underlying biological mechanisms.

To further explore this possibility we look at the correlation values (c.v.) (M 3.4) between the fingerprint parameters across all measurements as depicted in Figure 4de. We define an absolute c.v. below 0.3 as no (anti)correlation, between 0.3 and 0.6 as weak (anti)correlation and above 0.6 as strong (anti)correlation. While for most parameter pairs no correlation is observed, we find a strong dependency between parameters B, β and α and a weak anticorrelation between A and α. While α and β are positively correlated (0.81), both are negatively correlated with B (-0.81 and -0.92). Notably, the changes already observed for the more fluid-like HoxB8 cells do perfectly follow these correlations. A simultaneous variation of α, β and B might be a way for the cell to switch between a solid- and fluid-like state. This finding also implies that, even though the two power law approach is able to perfectly describe the observed viscoelastic properties, it might over-fit the data as some parameters are not fully independent. The cytoplasm does not seem to consist of two independent classes of materials but instead a deeper underlying mechanism might regulate the low-frequency elastic-like behaviour just as the high frequency more fluid-like behaviour.

In order to see if we can indeed further reduce the number of parameters required to describe the active intracellular mechanical state we performed a principal component analysis (PCA) (M 3.5). The explained variance ratio of each component is presented in Figure 4e left. Already 92% percent of the variance can be explained using only the first 3 components and 81% can even be explained with just 2 components. This supports the hypothesis that few underlying biological mechanisms act as “hyper parameters”, which tune the mechanical fingerprint of the intracellular space.

Each principal component is a superposition of the normalized fingerprint parameters. We therefore wondered which fingerprint parameters dominate the first 2 principal components. Looking at the relative contributions (Figure 4e right, SI 2.1), we find that the first component C1, that explains 62.09% of all variance in data, mainly consists of fingerprint parameters α, B and β (21%, 18%, 20%). Finding these parameters combined in a single principal component is expected as they also show strong correlation (Figure 4e) but strikingly, even though neither parameter by itself appeared to be dominant in characterizing the intracellular state in the previous analysis (Figure 4b,c), the combined component C1 seems to be the most relevant principal component. The second component C2 consists of the remaining fingerprint parameter E, γ and A (20%, 29%, 27%).

To determine whether C1 and C2 are already sufficient to distinguish between cell types, we plot the cell types according to these two parameters (Figure 4f). We observe that most cell types appear to separate from one another, indicating that C1 and C2 are effective in distinguishing between them. Interestingly, there are some cell types (HeLa, MDCK, A549) that cluster together, suggesting that they share similar mechanical properties. Whether this clustering is related to the fact that all these cells are of epithelial origin remains to be studied in more detail in future work.

As the first 2 components are sufficient to differentiate between most different cell types we wondered if they can be mapped on a physical interpretation. The first component C1 consists mainly of the parameters α, β and B. Additionally to the high correlation between the parameters, the increase of C1 represents an effective switch from a more solid material towards a fluid material, which we interpret hence as a parameter marking the fluidity of the cell. In contrast, the second component is largely a combination is solid-like material prefactor A and the intracellular activity parameter E. Hence, increasing C2 may either be interpreted as cellular stiffening or increase in mechanical activity. To avoid this ambiguity, while conserving the key features of the principle component analysis we propose a qualitative three dimensional phase space which consists of activity (mainly E) cellular resistance (mainly A) and fluidity (principal component C1)(Figure 4g). Although in principle both, A and B contribute to the mechanical resistance of a material when moving thought it, our finding that A is always much larger than B motivates to identify A with a mechanical resistance. Observation of the different cell types in this face space allows to identify key functional differences (Figure 4 g). Although HoxB8 macrophages and cancerous A549 cells display the highest activity, they together with MCF7 cells show the most fluid like phase and the lowest resistance. Speculatively, a high activity paired with a decreased resistance and higher fluidity might be advantageous for both motile cancer but also immune cells. On the other hand C2C12 muscle cells show the highest resistance paired with the most solid like phase, underlying that these cells might have fundamentally different active mechanical properties then other cell types with different function. As already seen in the PCA, epithelial cell types HeLa, MDCK and MCF7 show overall more similar properties with the largest differences observed along the fluid-solid like phase.

Conclusion

The here introduced mechanical fingerprint now opens the way for a systematic and functional study and characterization of the intracellular mechanical properties. The finding that HeLa cells can largely compensate for the mechanical loss of support upon microtubule polymerization suggests a functional role for the know up-regulation of actin polymerization upon MT depolymerization. This notion that the mechanical fingerprint is tuned to the cellular function is further supported by the expected mechanical properties when extending the comparison to muscle and immune cells, where the increase in stiffness of muscle and decrease in viscosity on immune cells does perfectly fit their biological function. A key mechanical switch is a change between solid to fluid characteristics, which is directly reflected in a superposition of the fingerprint parameters that carries the most separating potential. This first principal component can be identified with a fluidity and completes the three dimensional phase space that we propose to use for placing cells in a functional relation to their mechanical fingerprint.

Material and Methods

Optical setup

In this study we use an optical tweezers setup that has been previously described34 to perform active and passive microrheology measurements and apply well defined forces onto phagocytozed probe particles in living cells. Briefly, the setup is based on a home-build bright field microscope equipped with an 60x NA 1.2 Objective, where we employ two different lasers. A 808 nm laser (LU0808M250, 808 nm, 250 mW Lumics GmbH, Berlin,331Germany) is operated at moderate power (70 mW at sample plane) to allow for stable optical trapping and force application to probe particles at the focal plane. The lateral position of this laser can be precisely controlled with a piezo-based tilting mirror. Using a high NA=1.4 condenser, most of the transmitted laser light is collected. By projecting the back focal plane of the condenser onto a position sensitive detector (PSD), the trapping forces can be directly quantified.42 A second, stationary, laser (L976-PAG500, 976 nm, 500 mW, Thorlabs, New Jersey, USA) is operated at low laser power (1 mW at sample plane) to not apply any relevant trapping forces. This laser is utilized to monitor the bead displacement via back focal plane interferometry43 with nm precision. A schematic of this setup is depicted in Figure 1a.

Active and passive microrheology in cells

In order to conduct active and passive microrheology measurements, we utilized the previously desribed optical tweezers setup (see Figure 1a). Cells were prepared following the protocols outlined in Materials and Methods M 4. Throughout the experiments, the cells were maintained at 37 °C and 5% CO2 to ensure physiological conditions. We investigated a minimum of N=60 cells for each cell type or condition over the course of n=3 consecutive days. Data points were recorded at a sampling rate of 65 536 Hz.

Active microrheology

In active microrheology, the 808 nm trapping laser is employed to oscillate with a variable driving frequency fD, exerting a sinusoidal force F (t) onto a probe particle located inside a cell. Simultaneously, a second position detection laser records the position x(t) of the particle. The relationship between the particle’s position and the applied force is determined by the response function χ(t), as given by: In the frequency domain, this convolution can be evaluated as a product, providing access to the response function: which characterizes the viscoelastic properties of the probed region. Here, and denote the Fourier transform of position and force.

To obtain the frequency-dependent response function, the probe particle is successively driven at different driving frequencies (fD ∈ 1Hz, 2Hz, 4Hz, 8Hz, 16Hz, 32Hz, 64Hz, 128Hz, 256Hz, 512Hz, 1024Hz). At each frequency, a minimum of 3 periods were recorded, with a maximum duration of 1 second. Utilizing the generalized Stokes-Einstein relation:44 where R represents the particle radius, we determine the complex shear modulus G∗(f).

Passive microrheology

In passive micro rheology measurements, free particle fluctuations inside the cells are recorded without application of external forces. In this study we employ the weak detection laser and back focal plane interferometry to track the bead displacement with high spatial and temporal precision. The particle position x(t) is tracked 3 consecutive times over a time span of 10 s for each cell investigated. For each time series x(t), we calculate the power spectral density (PSD) and average it over the 3 repetitions.

In equilibrium systems, the PSD is linked to the mechanical properties of the system via a generalized fluctuation-dissipation theorem (FDT):44, 45 Here kBT is the thermal energy of the system and χ′′(f) the imaginary part of the response function. In passive systems, this equation is valid and can be used to determine the mechanical properties of the system. Living cells, however, are non-equilibrium systems. Metabolic energy leads to stronger particle fluctuations than one would expect for a passive systems. We define an effective energy by adding an active cellular energy Eactive to the thermal energy: This is the energy required to explain the particle fluctuations given the mechanical properties χ′′(f) measured in active microrheology. This yields a new adapted version of the FDT: By separately measuring χ′′(f) in active microrheology and in passive microrhelogy, we gain access to the effective energy of the system and thereby the level of non-equilibrium: The effective energy quantifies the level of activity in the system. Any deviation from 1 is indicative for non-equilibrium.

Statistical analysis

Bootstrapping

Bootstrapping is employed to get an estimate for the distribution of the mean cellular active mechanical properties. The following scheme is used:

We have i experimental conditions, either different cell-types or drug treatments. For each condition i, we investigate Ni different cells. For each cell, we obtain curves for G′,G′′ and EEff

For each condition i, the bootstrapping data set is generated. Ni samples are drawn with replacement with consecutive mean calculations. This process is repeated Nbootstrap=10.000 times for each quantity of interest.

The resulting distribution of curves captures the distribution of the means.

Each curve is subsequently fit with the respective model which in turn also yields the distribution of the mean fit parameters.

Significance tests

Statistical tests between parameters obtained by the bootstrapping scheme (M 3.1) are conducted in the following way, a visualization of this process is depicted in SI Figure 1.

We subtract the individual parameter distributions from each other. With this new distribution, we can check if 0 falls outside of the confidence interval.

We agree on confidence intervals for significance α = 0.05 for *, α = 0.01 for ** and α = 0.001 for ***.

Next we do a Bonferroni correction: αcorr = α/i with i being the number of experimental conditions.

If 0 is outside the interval [αcorr : 1 − αcorr], then the significance level is reached

Z-score

We use an adapted z-score to quantify how different a certain parameter is between two different cell types. Here σ is the standard deviation SD.

Correlation analysis

Correlation analysis was performed using the corrcoef function of the python package numpy.46

Principal component analysis PCA

In order to see if the complexity of the 6 parameter fingerprint E0, E1, A, α, B, β can be further reduced, we performed a principal component analysis on the data. To this end, we utilized the function sklearn.decomposition.PCA from the Python module skicit learn.47The data was normalized prior to the pca using the sklearn.preprocessing.StandardScaler method.

Cell culture and bead insertion

A549, C2C12, CT26, HeLa, and MDCK cells were cultured in Dulbecco’s modified Eagle medium (DMEM, Capricorn) supplemented with 1% Penicillin Streptomycin (Gibco) and 10% fetal bovine serum (FBS, Sigma-Aldrich) at a temperature of 37°C and 5% CO2.

HoxB8 cells, provided by the Institute for Immunology at the University of Münster, Germany, were cultured according to a protocol described elsewhere.11 To induce differentiation into macrophages, HoxB8 cells were suspended in differentiation medium three days prior to the experiment. After this, the same procedure as for the other cell lines, with the exception of using EDTA instead of Trypsin, was applied.

Cells were split when they reached near full confluency and seeded onto Fibronectin-coated glass cover slips (22x55x0.15 mm VWR). Medium was then aspirated, and a fresh 1:10,000 dilution of 1 µm beads (Polybead® Microspheres 1 µm, Polyscience, Inc) in medium was added to the sample . Cells where incubated for up to 15 hours to allow for adequate bead phagocytosis. Then, cells were washed with PBS to remove any residual extracellular beads. Following aspiration of PBS, a cover slip was fixed to the estimated cell spreading area using two layers of 200 µm adhesive tape (DST1950, Thorlabs, New Jersey, USA).

The sample chamber was filled with CO2-independent medium (CO2 Independent Medium, Gibco™) to ensure optimal cell viability during the experiments.

Pharmacological treatment

Cytochalasin B (Sigma-Aldrich), used for depolymerization of actin filaments, and Nocodazole (Sigma-Aldrich), employed for microtubule depolymerization, were employed in this study. Both drugs were applied at a concentration of 10 µg mL−1 by adding them to the medium 10 minutes prior to the experiment.

eLife Assessment

Münker and colleagues use an optical tweezer setup to apply oscillatory forces to endocytosed/phagocytosed glass beads over a wide frequency range (from ~1 to 1000 Hz) and probe cytoplasmic material properties at multiple time scales in six different cell types. Using statistical methods and principal component analysis, they find that the active and passive mechanical properties of cells can be described by 6 parameters (from power law fits) that allow characterizing the viscous and elastic nature of the cytoplasmic material as well as an effective active energy driven by cellular metabolism. Overall, this is very well done and important work, using convincing and state-of-the-art methods, albeit with some limitations related to the way the beads are internalized.

Reviewer #1 (Public Review):

Summary:

In this MS, Muenker and colleagues, explore the intracellular mechanics of a range of animal adherent cells. The study is based on the use of an optical tweezer set up, which allows to apply oscillatory forces on endocytosed/phagocytosed glass beads with a large frequency range (from ~1 to 1000 Hz) , allowing to probe cytoplasm material properties at multiple time scales. By switching off the laser trap, the authors also record the positional fluctuations of beads, to extract passive rheological signatures. The combination of both methods allow to fit 6 parameters (from power law fits) that allow to characterize the viscous and elastic nature of the cytoplasm material as well as an effective active energy driven by cellular metabolism. Using these methodologies, the authors first establish/confirm, using HeLa cells, that the cytoplasm is more solid like at short frequencies, and more fluid like at higher frequencies, and that these material states depend on both microtubules and actin cytoskeleton. The manuscript then go on to explore how these parameters evolve in other 6 cell types including muscles, highly migratory and epithelial cells. These results show for instance that muscle cells are much stiffer, while migratory cells are more fluid like with an increased active energy. Finally using statistical methods and principal component analysis, the authors establish some mechanical fingerprints (activity, fluidity and resistance) that allow to distinguish cell's mechanical state and relate it to their particular functions.

Strengths:

Overall this is a very well-executed work, which provides a large body of rigorous numbers and data to understand the regulation of cytoplasm mechanics and its relation to cell state/function.

Weaknesses:

A limit of the paper is that the biological mechanisms by which intracellular mechanics is modulated (e.g. among cell types) remains unexplored and only briefly discussed. Yet this limit is greatly offset by the rigor of the approach.

Reviewer #2 (Public Review):

Summary:

By analyzing cells' frequency-dependent viscoelastic properties and intracellular activity through microrheology, Münker et al simplify the complex active mechanical state into six key parameters that constitute the mechanical fingerprint. They apply this concept to cells treated with cytoskeleton-inhibiting drugs. Additionally, a comprehensive statistical analysis across various cell types shows how cells coordinate their mechanical properties within a defined phase-space marked by activity, mechanical resistance, and fluidity.

Strengths:

(1) The distribution of the six parameters: they have been well characterized based on established theories, and they can be used to understand cell-type-specific biomechanical differences. The examples of muscle cells and immune cells were profound and informative.

(2) Efforts to perform dimension reduction of parameter space into activity (E), fluidity (C1) and resistance (A) are insightful and will be helpful for future characterization of cell mechanics.

Weaknesses:

(1) The most difficult part of the method is the part with actin polymerization inhibition with cytochalasin B. The data shows that viscoelastic parameters as well as active energy parameters are unaffected by cytochalasin B. It is reasonable to expect that elasticity will reduce and fluidity will increase upon application of such a drug. The stiffness-reducing effect was observed only when CB was used with nocodazole most likely because of phagocytosis of the bead, which is governed by microtubule. The use of other actin-depolymerizing drugs such as latrunculin A would be needed to test actin's role in mechanical fingerprints. If actin's role is only explained by accompanying microtubule inhibition, it is not a convenient system to directly test the mechano-adaptation process.

(2) Depolymerization of MT with nocodazole did not reduce the solid-like property A. Adding discussion and comparison with other papers in the literature using nocodazole will be helpful in understanding why.

(3) Overall, the usefulness of the concept of mechanical fingerprints and comparisons with other cell mechanics studies (from other groups) will make this manuscript stronger.

Reviewer #3 (Public Review):

Summary:

Cells and tissues are viscoelastic materials. However, metabolic processes that underly survival, growth and migration render the cell as an active matter at non-equilibrium. These two facts contribute to the difficulty of probing mechanical properties especially with sub-cellular resolution. However, the concept that the mechanical phenotype can be indicative of normal physiology necessitates approaches of defining the cellular phenotype. Here, Muenker et al evokes a powerful argument for mapping intracellular mechanics using optical tweezer- active microrheology. They present a suite of parameters towards a definition of a mechanical fingerprint. This is a compelling idea. There are some concerns as detailed below

Strengths:

These are technically challenging experiments and the authors provide systematic approaches to probe a system at non-equilibrium.

Weaknesses:

The importance of the mechanical fingerprint is diluted due to some missing controls needed for biological relevance.

Author response:

The following is the authors’ response to the original reviews.

General comments:

You will see that many of the reviewers’ comments overlap. From our discussion with them, we agree that several of these comments should be addressed in this study, particularly comments related to the interpretation of the effect of drugs acting on the cellular cytoskeleton (reviewers #1 and #2). We also agree that the comparison of isogenic cell lines such as the mcf10a series or the 4T1 series should address some of the concerns regarding the interpretation of the mechanical fingerprint (reviewer #3). Also, certain methodological aspects should be easily clarified (reviewers #1 and #3).

We also agreed that other comments may be more difficult to address in the context of this study. This is the case for comments related to establishing a link between different mechanical signatures and different cellular functions/outcomes (Reviewers #1 and #3). One could test whether migration or proliferation is altered by changing the mechanical fingerprint, or you could simply discuss these aspects by carefully reviewing the literature to corroborate mechanical signatures with known cellular phenotypes (e.g. migration speed, adhesion, cell size...). This is also the case for comments on the influence of other cellular parameters such as molecular crowding and energy metabolism, which could be left for future work or where you could use a low dose of cycloheximide (below the level of deleterious effects) to address the effect of cytoplasmic proteins (reviewer #3).

We thank the editor for providing this helpful overview of the requested revisions. We have carefully addressed these points throughout the revised manuscript. The only difficulty was to establish the isogenic cell lines as requested. It took us over 18 months to find a source of these cells in Europe, and since then we are trying hard, but not successful to get these cells stably growing in the condition necessary for the optical tweezers experiments. As we have now spent more than 2 years on this without success, we decided to resubmit the paper without this part to not further delay this manuscript. The additional experiments and revisions have substantially strengthened the manuscript. Especially, the addition of Latrunculin A as suggested was an excellent request, as now the results regarding actin depolymerization and mechanical properties are in excellent agreement with the expected effects, as Latrunculin A is much more efficient in depolymerizing actin than cytochalasin B. The major changes are summarized below, followed by a detailed point-by-point response to all reviewer comments.

General changes

(1) Repeated measurements on wild-type HeLa cells.

(2) Repeated all Cytochalasin B and Nocodazole experiments and increased the number of analyzed cells to approximately 60 per condition.

(3) Performed additional experiments using Latrunculin A and combined Latrunculin A + Nocodazole treatment.

(4) Performed immunostainings for all cytoskeletal perturbation conditions (WT, Cytochalasin B, Latrunculin A, Nocodazole, Cytochalasin B + Nocodazole, and Latrunculin A + Nocodazole).

(5) Refined the rheological analysis procedure and expanded the methodological description.

(6) Revised the manuscript text throughout and expanded the discussion of limitations and biological interpretation.

Public Reviews:

Reviewer #1 (Public Review):

A limit of the paper is that the biological mechanisms by which intracellular mechanics is modulated (e.g. among cell types) remains unexplored and only briefly discussed. Yet this limit is greatly offset by the rigor of the approach.

We thank the reviewer for this positive assessment and agree that a more extensive discussion of the biological mechanisms underlying the observed mechanical fingerprints strengthens the manuscript. We have substantially expanded the Discussion and Conclusion sections to address potential contributions of cytoskeletal organization, intracellular transport, molecular crowding, and metabolic state. In addition, we now discuss the relationship between the identified mechanical phase space and known cellular phenotypes where appropriate, while explicitly outlining the limitations of the current study and the need for future investigations linking intracellular mechanics to cellular function.

Reviewer #2 (Public Review):

The most difficult part of the method is the part with actin polymerization inhibition with cytochalasin B. The data shows that viscoelastic parameters as well as active energy parameters are unaffected by cytochalasin B. It is reasonable to expect that elasticity will reduce and fluidity will increase upon application of such a drug. The stiffness-reducing effect was observed only when CB was used with nocodazole most likely because of phagocytosis of the bead, which is governed by microtubule. The use of other actin-depolymerizing drugs such as latrunculin A would be needed to test actin’s role in mechanical fingerprints. If actin’s role is only explained by accompanying microtubule inhibition, it is not a convenient system to directly test the mechano-adaptation process.

We thank the reviewer for this important suggestion. To strengthen the interpretation of the actin perturbation experiments, we repeated the Cytochalasin B measurements with an increased number of cells and performed additional experiments using Latrunculin A, a mechanistically distinct and more potent actin-depolymerizing compound. Together with complementary immunostaining experiments, these additional data reveal distinct contributions of the two major cytoskeletal systems to the intracellular mechanical fingerprint. Whereas actin depolymerization primarily affects intracellular stiffness and fluidity, microtubule depolymerization has the strongest effect on intracellular activity while also contributing to cellular softening. These additional experiments provide a substantially clearer interpretation of the respective roles of actin filaments and microtubules in shaping the intracellular mechanical fingerprint.

Depolymerization of MT with nocodazole did not reduce the solid-like property A. Adding discussion and comparison with other papers in the literature using nocodazole will be helpful in understanding why.

We thank the reviewer for this suggestion. We have expanded the discussion and now compare our observations with previous AFM studies investigating Nocodazole treatment. While AFM measurements of cortical mechanics often report little change or even increased stiffness after microtubule depolymerization, our intracellular measurements reveal pronounced softening and strongly reduced intracellular activity. We now discuss that this difference likely reflects the distinct intracellular mechanical compartment probed by intracellular microrheology compared with cortical AFM measurements.

Overall, the usefulness of the concept of mechanical fingerprints and comparisons with other cell mechanics studies (from other groups) will make this manuscript stronger.

We thank the reviewer for this suggestion. Throughout the revised manuscript we have strengthened the comparison of the mechanical fingerprint with previous literature. In particular, we now discuss the cytoskeletal perturbation experiments in the context of published AFM studies, compare the observed mechanical differences between cell types with previous measurements where available, and expand the discussion of the biological interpretation and limitations of the proposed mechanical fingerprint.

Reviewer #3 (Public Review):

The importance of the mechanical fingerprint is diluted due to some missing controls needed for biological relevance.

We thank the reviewer for raising this important point. To strengthen the biological interpretation of the mechanical fingerprint, we performed substantial additional experiments, including repeated cytoskeletal perturbation measurements with increased sample sizes, additional Latrunculin A experiments, and complementary immunostaining analyses. We also expanded the discussion to address the influence of factors beyond the cytoskeleton, including molecular crowding and metabolic state, and explored possible relationships between the proposed mechanical phase space and cellular phenotypes. While we agree that future studies using well-controlled isogenic model systems will be required to establish direct links between intracellular mechanics and biological function, we believe that the additional experiments and expanded discussion substantially strengthen the biological relevance of the present study.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

A caveat of the general methodology, which is partially acknowledged in the MS is that beads are endocytosed and likely end up in specific lysosomal compartments. Therefore, it is not clear whether the mechanical fingerprint fully represent the material properties of bulk cytoplasm, and not something more specific to lysosomal organelles. For instance, lysosome motion may be largely driven by motors moving along MT cytoskeletal track, and the extracted effective energy may as such not fully represent the crowding and effective active temperature of the cytoplasm. This limit certainly affect the interpretation of the results in other cell types, in which membrane trafficking and cytoskeletal organization may vary largely. I believe it would be very important to outline this limitation of the work and discuss it in light of the results obtained throughout.

We thank the reviewer for pointing out this important limitation, which was not sufficiently addressed in the original manuscript. We have now acknowledged this issue throughout the manuscript and added a limitation section to the conclusion to clarify that our findings specifically relate to internalized objects surrounded by a membrane and therefore primarily reflect the properties of membrane-bound organelles in the 1 µm size regime, rather than the bulk cytoplasm as a whole.

We consider this focus on membrane-enclosed intracellular objects to be biologically relevant and interesting in its own right. Alternative approaches for introducing tracer particles, such as microinjection or particle guns, are generally more invasive and less reproducible. We therefore deliberately focused on phagocytosed beads as a minimally perturbative and robust experimental system in this study.

The evolution of the mechanics in Hela Cells using cytoskeletal drugs in interesting, but I was confused by the fact that authors interpret the effect of cytochalasin solely on the cortex. As they are probing intracellular rheology, variations (or lack thereof) may rather reflect bulk F-actin networks? Also the compensation mechanism is interested, but it would need to be strengthened by immunostaining for instance, to support the claim, that microtubule depolymerization enhances F-actin networks.

We thank the reviewer for this important comment. To elaborate on the effect of cytoskeletal filaments, we extended our analysis by repeating the experiments, increasing the number of samples, and investigating the effect of an additional drug, Latrunculin A. Additionally, we conducted immunostaining with subsequent confocal imaging to deepen our understanding of the effect of the respective drugs. The additional experiments reveal that actin and microtubules contribute differently to the fingerprint. Actin depolymerization primarily affects intracellular stiffness and fluidity, whereas microtubule depolymerization has the strongest effect on both mechanics and intracellular activity. Combined perturbation produces the largest overall effect. Based on these additional data, we no longer invoke the compensation mechanism proposed in the original manuscript. While interactions between the actin and microtubule cytoskeleton have been reported previously, our immunostaining experiments do not provide evidence for a compensatory increase in actin organization following microtubule depolymerization. We have therefore removed this interpretation from the revised manuscript and replaced it with a discussion based on the newly acquired perturbation and imaging data.

The final figure using principal component analysis is very interesting, but it would be important to link this to phenotypic signatures of the different cells. Could the authors try to link resistance, fluidity and activity to the different functions/behavior of cells? For instance, some of these cells are migratory but some may move much faster than others, and it would be very interesting to correlate the degree of activity or fluidity with speed of migration, or cell shape/size/contractile state for example.

Indeed, this is an important point. Establishing direct links between the mechanical fingerprint and functional cellular properties such as migration, contractility, proliferation, or morphology would substantially strengthen the biological interpretation of the identified phase space. We carefully considered this suggestion and explored several approaches to relate the measured mechanical parameters to cellular phenotype. However, obtaining directly comparable quantitative functional data across all investigated cell types proved challenging. Parameters such as migration speed, adhesion, and contractility depend strongly on experimental conditions, including substrate properties, assay design, and culture conditions, making literature values difficult to compare across studies. To address the reviewer’s concern, we expanded the discussion and incorporated comparisons to available literature where appropriate. For example, previous studies have reported higher migration rates for HeLa cells compared with MCF7 cells, which is qualitatively consistent with the higher intracellular activity observed in HeLa cells. However, due to the limited comparability and availability of quantitative functional data across the investigated cell types, we refrained from performing a formal correlation analysis. In addition, we grouped the investigated cell lines according to several broad phenotypic classifications, including epithelial/mesenchymal character, cancer status, metastatic potential, and migratory potential, and examined their distribution within the proposed phase space. While this exploratory analysis provides additional biological context, it did not reveal robust relationships that could support definitive conclusions regarding structure–function relationships. We therefore agree with the reviewer that establishing direct links between intracellular mechanical fingerprints and cellular function represents an important next step. To this end, future studies will combine intracellular rheological measurements with independently quantified functional assays, ideally in well-controlled isogenic model systems.

Reviewer #2 (Recommendations For The Authors):

The study needs more thorough validation against known technology (such as AFM) or literature, e.g., rheological change upon the same drugs used in the current study.

We thank the reviewer for this suggestion. We have expanded the discussion of the cytoskeletal perturbation experiments and now compare our observations to previous AFM studies and related literature on cytoskeletal mechanics. Consistent with AFM measurements of cortical mechanics, actin depolymerization using Cytochalasin B or Latrunculin A resulted in a reduction of cellular stiffness. In contrast, microtubule depolymerization produced effects that differ from many AFM studies, which report either no change or an increase in cortical stiffness following Nocodazole treatment. We now explicitly discuss that this discrepancy likely reflects the different mechanical compartments probed by the two techniques. AFM predominantly measures the actin-rich cell cortex, whereas our intracellular microrheology measurements probe the mechanical environment experienced by membrane-bound intracellular particles. We therefore interpret the differing response to microtubule depolymerization as evidence that intracellular active mechanics and cortical mechanics can be influenced by distinct physical mechanisms. These comparisons have been incorporated into the Results and Discussion sections of the revised manuscript.

Page 8: Citation to Fig. 3a is missing before mentioning Fig. 3b.

We revised the manuscript to ensure that all references are given in an appropriate order.

Proper uses of hyphens are recommended to avoid confusion. For example, ’a yet not understood change’ can be written as ’ a yet-not-understood change’.

We thank the reviewer for this suggestion. We carefully revised the manuscript to improve the use of hyphenation and compound modifiers throughout the text. The specific example highlighted by the reviewer, as well as similar constructions, have been corrected to improve readability and avoid ambiguity.

Reviewer #3 (Recommendations For The Authors):

As it reads, sinusoidal waves are applied sequentially from 1- 1024Hz. Please clarify if amplitude is the same for each frequency, also how many frequencies are used? On this point, due to perturbations due to alterations in pre-stress, are the orders of frequencies randomized?

We thank the reviewer for pointing out this ambiguity. We have revised the manuscript to provide a more detailed description of the active microrheology protocol. Specifically, we now state that all measurements were performed using a constant trapping-laser oscillation amplitude of 200 nm and that the applied frequencies were 1, 2, 4, 8, 16, 32, 64, 128, 256, 512, and 1024 Hz. The frequencies were applied sequentially in increasing order and were not randomized. This information has now been added to the manuscript.

How many beads are probed in a given cell?

We thank the reviewer for this question. We have clarified this point in the Methods section and now explicitly state that only a single phagocytosed probe particle was analyzed per cell. Of course, many different cells, and hence beads, have been analyzed per cell type.

Is the graph in 1 c G’, G” per cell or average of many cells?

We thank the reviewer for pointing out this ambiguity. In the original version of the manuscript, Figure 1b showed data from a representative cell, whereas Figure 1c displayed an average over multiple cells. To avoid confusion, we revised Figure 1 and now show representative data from a single measurement throughout the analysis workflow (Figure 1c,e,f).

Figure 1e is quite nice, however, is there an equivalent performed in a nonlinear ECM such as collagen for comparison, in a similar vein can the equivalent be calculated for cells with/without treatment with low doses of cycloheximide to reduce protein synthesis? Yes, cytoskeletal elements are important for cell mechanics, but cytoplasm crowding is often an overlooked factor.

We thank the reviewer for this important suggestion, and we are glad that the reviewer likes figure 1e. Regarding non-linear ECM, we have not done such experiments using optical tweezers. Collagen is a highly heterogeneous material and using the small deformations that we can obtain using the optical tweezers, our access to the non-linear contributions is rather limited.

However, we agree that factors beyond the cytoskeleton, including molecular crowding and protein content, can make important contributions to intracellular mechanics. While investigating these effects experimentally, for example through cycloheximide treatment, would be highly interesting, such studies were beyond the scope of the present work.

The primary focus of this study was to establish and validate a mechanical fingerprint for intracellular active microrheology and to investigate how this fingerprint responds to perturbations of the cytoskeleton. The additional experiments performed during revision therefore concentrated on strengthening the interpretation of the cytoskeletal contributions.

At the same time, we agree that molecular crowding represents an important alternative mechanism influencing intracellular mechanics. We have therefore expanded the Discussion and Conclusion sections to explicitly acknowledge this limitation and now cite recent studies demonstrating strong effects of molecular crowding on intracellular rheology (Umeda et al,. 2023, Ebata et al., 2023). We further discuss that, besides cytoskeletal organization, metabolic state, intracellular transport, and molecular crowding are likely contributors to the observed mechanical fingerprint.

The biggest issue is the interpretation of the different factors as each of these cells have different energetic needs. The comparison between cancer cells with different aggressiveness, immune and epithelial cells. For example, some types of cancer cells will be dominated by glycolysis vs oxphos, which will influence both the cytoplasmic and nuclear mechanics? It would be useful to carefully assess factors not restricted to

(a) Cytoskeleton

(b) Protein synthesis

(c) Metabolic state

For similar lines and/or cells where there are lineages that are either more metastatic in cancer, normal counterpart or drug resistant in an effort to link the fingerprint to a biological output. Specifically, is migration, proliferation, survival correlated with the measurements. The reviewer is sensitive to the technical difficulties of the experiments. However, the interpretation and importance of the mechanical fingerprinting requires additional work as mentioned above.

We thank the reviewer for this thoughtful comment. We agree that intracellular mechanics is likely influenced by a broad range of biological factors beyond the cytoskeleton, including metabolic state, molecular crowding, intracellular transport processes, and protein synthesis. We also agree that the biological significance of the mechanical fingerprint would be strengthened by establishing direct links to functional cellular outputs such as migration, proliferation, or survival. To address the first point, we have expanded the Discussion and Conclusion sections of the manuscript to explicitly acknowledge that the observed fingerprint is unlikely to be determined solely by cytoskeletal organization. In particular, we now discuss the potential contributions of metabolic state, intracellular transport, and molecular crowding, and cite recent studies demonstrating the importance of these factors for intracellular mechanics. To address the second point, we explored several strategies to relate the measured mechanical fingerprints to cellular phenotype. We expanded the discussion of available literature, including examples where mechanical properties and migratory behavior appear qualitatively consistent. In addition, we grouped the investigated cell lines according to broad biological characteristics, including epithelial/mesenchymal character, cancer status, metastatic potential, and migratory potential, and examined their distribution within the proposed phase space. While this exploratory analysis provides additional biological context, it did not reveal robust relationships that would support definitive conclusions regarding structure–function relationships. We therefore agree that establishing direct links between intracellular mechanics and cellular function represents an important next step. Such studies will require quantitative functional assays performed under controlled and directly comparable conditions, ideally using well-defined isogenic model systems. We now discuss these limitations and future directions explicitly in the revised manuscript.

Version 2

Abstract

Living cells are complex entities that perform various tasks with astonishing robustness. While the direct dependence of biological processes on controlled protein expression is well established, we have only begun to understand how intracellular mechanical characteristics guide and support biological function. This is in stark contrast to the expected functional role that intracellular mechanical properties should have for many core cellular functions such as organization, homeostasis, and transport. From a mechanical point of view, cells are complex viscoelastic materials that are continuously driven out of thermodynamic equilibrium, which makes both a physical measurement and mathematical modeling of their properties difficult. Here, we define a “mechanical fingerprint” that can not only characterize the intracellular mechanical state, but also carve out the mechanical differences between cell types with the potential to relate these to proper cell function. By analyzing the frequency-dependent viscoelastic properties and intracellular activity of cells using microrheology, we distilled the complex active mechanical state into just 6 parameters that comprise the mechanical fingerprint. The systematic investigation of the fingerprint illustrates a parameter tuning that could be related to functional cellular requirements. However, the full potential of the mechanical fingerprint is given by a statistical analysis of its parameters across all investigated cell types, which suggests that cells can modify mechanical parameters in a correlated way to position their intracellular mechanical properties within a well-defined phase-space that is spanned between activity, mechanical resistance, and fluidity. This paves the way for a systematic study of the interdependence of biological function and intracellular active mechanics.

Main

Introduction

Cells are highly complex, heterogeneous materials that tightly control their viscoelastic characteristics by precise tuning of the active mechanical properties to maintain a non-equilibrium mechanical steady state. This mechanical homeostasis requires a continuous consumption of metabolic energy, mostly in the form of ATP, and is maintained while the system undergoes various dynamic functions ranging from tissue morphogenesis and patterning1, 2 to organelle positioning,3 differentiation,4 migration,5, 6 and polarization.7

Although it is well-acknowledged that intracellular mechanical properties are key to understanding major cellular functions, most studies have measured cellular mechanics from outside of the cell using e.g. AFM,8, 9 micropipette aspiration,10 traction force microscopy,11 tether pulling,12 magnetic twisting cytometry,13 or deformability cytometry.14 While these studies allowed deep insights into the relevance of cell mechanics for supracellular biological processes by probing predominantly the stiff actin cortex that surrounds the cell, inferring information about the mechanical properties of the cytoplasm remained difficult.15 This experimental gap has recently been bridged by active intracellular rheology experiments that are based on optical as well as magnetic tweezers.16–19 For example, a recent study has shown that the cytoplasm softens during mitosis,20 which might be key for the proper functioning of this process.

While a large body of work has focused on the viscoelastic properties of cells, it has become increasingly clear that the well-controlled active force generation of living cells has to be integrated to obtain a full intracellular mechanical description. Recent studies have therefore focused on investigating these non-equilibrium, active properties of cells.21–23 These active, energy-consuming cellular processes are a key component to maintain the intracellular organization despite the ever-dispersing forces of entropy.16, 24 While intracellular transport continuously organizes membrane and organelle architecture via motor activity, the dynamic growth and disassembly of cytoskeletal filaments ensures global processes such as migration,25 proliferation,26 or wound healing.27 Further, recent studies have shown that dynamic waves of actin filaments also organize mitochondria position during cell division28 and that cytoplasmic forces can reorganize nuclear condensates,29 thus emphasizing the key importance of active forces for fundamental cellular function.

Despite the generally accepted relevance of intracellular active mechanics for the functioning of entire organisms, it remains difficult to quantify and even harder to model theoretically. From a physical perspective, cells behave as viscoelastic materials where timescales are key information to correctly identify the dominating mechanical principles. Therefore, a full quantitative description requires microrheological information, typically provided in the form of a frequency-dependent, complex shear modulus G*(f).30 Although this quantification of the viscoelasticity is experi-mentally accessible, its interpretation requires theoretical models to become physically meaningful and to reduce the complexity of the data. However, while simple mechanical analogies, such as Kelvin-Voigt or Maxwell models lack the complexity to describe the observed behavior, more elaborate models, such as chains of springs and dashpots, require many parameters and thus do not reduce the complexity significantly. This has led to double power-law approaches, which are motivated by fractional Generalized Kelvin-Voigt (fGKV) models.31 In recent work, these have been successfully used to quantify the viscoelastic properties of fibroblasts,32 endothelial,33 and epithelial cells20 and they reduce the complex mechanical behavior to just four parameters. Recent efforts to include active force generation in a mechanical characterization were able to describe the active energy injected into the cellular system, also by a power-law20 that is defined by two additional parameters.

These insights raise the question to what extent living cells actively control their intracellular mechanical properties, or if these are simply byproducts of the cytoplasmic composition. Assuming a functional relevance, the parameters may vary considerably between cells and might even allow to identify different cell types and cellular situations in a fingerprint-like fashion. The relevance of this 6 parameter description was recently established by showing systematic changes during cell division, and to perform new statistical analysis of stochastic particle trajectories.20, 34 –36 In this study, we use optical tweezers-based active microrheology and passive observation of intracellular particle motion to demonstrate that cell types can be successfully distinguished in a pairwise comparison to position cells on 3D phasespace that reflects activity, rigidity, and fluidity.

Results and Discussion

Mechanical parameter set quantifies active mechanics

Intracellular viscoelastic material properties are commonly quantified by a frequency-dependent, complex shear modulus that we directly determine via optical tweezers-based active microrheology (M 1). Here, a custom-built optical tweezers setup (Figure 1a, M 2) exerts well-defined oscillatory forces onto phagocytosed 1 µm sized probe particles inside cervical cancer cells (HeLa). In this study, probe particles were added to the cell culture medium and taken up by the cells via endocytosis. Thus, they were surrounded by a membrane, and the obtained mechanical properties may serve as a proxy for membrane-bound organelles of similar size. To determine the resulting bead displacement, a stationary, weak detection laser monitors the particle position with nanometer precision. Applying sinusoidal forces (Figure 1b) at increasing frequencies of 1, 2, 4, 8, 16, 32, 64, 128, 256, 512, and 1024 Hz, while maintaining a constant trapping-laser oscillation amplitude of 200 nm, allows determination of the mechanical response function χ(f). This response function is then used to calculate the complex shear modulus according to G*(f) = 1/(6πRχ(f)), where R denotes the radius of the probe particle (Figure 1c). G*(f) is a complex quantity where the real part G′(f) (storage modulus) describes the elastic properties of the cell and the imaginary part G″(f) (loss modulus) describes the viscous properties (Figure 1c).

a) Viscoelastic properties and activity are measured within the cytoplasm of cells using a custom-built optical tweezers setup. b) Sinusoidal forces (blue) at varying frequencies are applied to a probe particle while the particle displacement (red) in response to this force is monitored. c) Repeating this procedure at different frequencies allows to calculate viscoelastic material properties in terms of the complex shear modulus G*(f) which consists of the storage modulus (blue) and a loss modulus (red). Both storage and loss modulus can collectively be fitted using a generalized fractional Kelvin-Voigt (fGKV) model (inset). d) This fGKV model consists of two power-laws, which in a simplified way can be pictured as a more solid-like material class resembling polymeric filaments and a more fluid-like material class (crowded molecules). e) Using an additional passive measurement, violation of the fluctuation-dissipation theorem can be directly visualized. (red area) f) The effective energy quantifies intracellular activity and can be fitted with a two-parameter power-law. g) Complex and frequency-dependent intracellular active mechanical properties can be reduced to a fingerprint of just 6 parameters, which describe the intracellular mechanical state.

To model the shear modulus, we follow the naive hypothesis that the cytoplasm consists of two main material classes, namely an elastic solid-like material that is generated by a sparse, connected polymer network, and a more liquid-like material that fills the bulk space of the cytoplasm (Figure 1d). Generally, such a combination of two complex materials can be approached by a linear combination of two springpots that leads to a double power-law model consistent with classical rheology data20, 31 –33: (1) G∗(f)=A(2πif)α+B(2πif)β

This complex function can then be used to simultaneously fit both the real and imaginary parts of the experimentally obtained shear modulus to determine the prefactors and power-law exponents. As shown in Figure 1c the model fits the experimental data well. R2-values of the fits are given for raw data (Appendix Table 16) and bootstrapped data (Appendix Table 15). Consistent with the initial hypothesis, we identify a region with a power-law exponent between 0 and 0.5 that we attribute to a more solid-like material (ideal solid has exponent 0), and a second power-law that is between 0.5 and 1 (ideal fluid has exponent 1) corresponding to a liquid-like material (Figure 1, right).

In addition to intracellular mechanics, the generated active forces are another key determinant of the intracellular state. As cells consume metabolic energy to drive motor proteins and many other cellular functions, cells are far from thermodynamic equilibrium. Experimentally, this is directly reflected in the motion of tracer particles within the cell. In thermodynamic equilibrium, the random motion of particles depends only on material properties and temperature, which is manifested by the fluctuation-dissipation theorem:37 (2) C(f)=kBTχ′′(f)πf

Here, C(f) is the power spectral density, quantifying the spontaneous fluctuations of the particle, χ(f) is the mechanical response function, and kBT is the thermal energy (M 1.2). In active, living cells, particle mobility is beyond motion explainable by thermal fluctuations38 (Figure 1e), as cytoskeletal rearrangement and transport processes also contribute to the motion of probe particles. An elegant way to summarize all active and thermal forces is by introducing an effective energy EEff 37(M 1). Intuitively, this effective energy reflects the ratio between observed particle motion C(f) and thermodynamically predicted motion that reflects the dissipative material properties as quantified by χ″(f): (3) EEff(f)=C(f)χ′′(f)πfkBT

It should be noted that this idea of an effective energy implies that the active forces generated by the cell share statistical characteristics with thermal forces. In this sense, we use the effective energy rather as a phenomenological quantity that measures the extent of non-equilibrium mechanical driving of the system. The free particle fluctuation spectrum C(f) is acquired by turning off the trapping laser while using the lower-power detection laser to monitor the particle position. The resulting effective energy measured for a representative HeLa cell (Figure 1f) shows an increased influence of active forces in the low-frequency regime, while the effective energy converges to the thermal limit of 1 kBT for higher frequencies. This directly demonstrates that metabolic energy dominates particle motion at long timescales, while assuming thermodynamic equilibrium at shorter timescales remains a valid approximation. This asymptotic behavior at short time scales is to be expected as motor proteins operate at longer time scales,39 and has previously been shown experimentally.16, 40 –42 This behavior was also used for an additional calibration method during optical tweezers experiments as described in Appendix 1.1. Interestingly, this suggests that the question about non-equilibrium processes can be reduced to a question of timescales, which is a potential explanation why this subject is so controversially discussed in literature.

Similar to the mechanical properties, the effective energy can also be fit by a power-law approach (Figure 1f): (4) EEff(f)=Efγ+1

This allowed us to describe the active forces generated by metabolic processes within the cell using only two parameters. Together with the four parameters describing the viscoelastic response, these quantities define a set of six parameters that characterize the intracellular mechanical state of a cell, as shown for the active mechanical properties of the cytoplasm in HeLa cells in Figure 1f. As we show in the following, the set of these 6 parameters can be interpreted as a mechanical fingerprint of a cell (Figure 1g).

In this study, we explore the variation of the mean active mechanical quantities G′, G″, and EEff across different cell types or conditions. To obtain a robust and reliable distribution of these curves, we apply a bootstrapping scheme to the recorded single-cell data (see M 3.1). The bootstrapped data is then subjected to respective model fitting, providing estimations for the distributions of the mean fingerprint parameters for each condition. Unless otherwise specified, all presented data is based on the bootstrapped data.

The cytoskeleton selectively affects the mechanical fingerprint

After establishing that the mechanical fingerprint provides an adequate description of the intracellular mechanical state of HeLa cells, we investigated whether individual fingerprint parameters can be selectively modulated by pharmacological perturbation of the cytoskeleton (Figure 2a). Cellular mechanics are commonly attributed to the actin cytoskeleton, which forms the cell cortex and a variety of intracellular structures. To assess the contribution of actin, we treated HeLa cells with 10 µg/mL Cytochalasin B (CytoB or CB) and 1 µM Latrunculin A (LatA), two compounds that interfere with actin polymerization through distinct mechanisms (Figure 2a).

Effect of cytoskeletal perturbations on the intracellular mechanical fingerprint of HeLa cells.

a) Composite images of HeLa WT cells treated with Cytochalasin B (CytoB), Latrunculin A (LatA), Nocodazole (Noc), Cytochalasin B + Nocodazole (CytoB+Noc), and Latrunculin A + Nocodazole (LatA+Noc). DNA stained with Hoechst in blue, microtubules stained with α-tubulin antibody in red, actin stained with phalloidin in green, and probe particles in white. Scale bar: 20 µm b-j) Changes in the fingerprint parameters and corresponding rheological quantities following cytoskeletal perturbation. The dashed lines in plots for G′, G″ and EEff show the models depicted in Figure 1c,f using the mean fit parameters A, B, E, α, β and γ.

CytoB treatment resulted in a modest but statistically non-significant decrease of the solid-like prefactor A (p = 0.08; Figure 2b), suggesting only a minor reduction in intracellular elasticity. Consistently, both the storage and loss moduli were slightly reduced, particularly at low frequencies (Figure 2d,g). LatA treatment produced a qualitatively similar but substantially stronger response. The solid-like prefactor A was significantly reduced (Figure 2b), while the power-law exponent α increased (Figure 2c), indicating a softer and more fluid-like intracellular environment. These observations are qualitatively consistent with previous AFM studies that reported a reduction in cortical stiffness following actin depolymerization.43, 44 In addition, both actin-disrupting treatments significantly reduced intracellular activity, as reflected by a decrease in the effective-energy prefactor E (Figure 2h,j). The more pronounced effect of LatA is consistent with its mechanism of action. While CytoB primarily caps filament ends and therefore leaves stabilized actin fila-ments largely intact, LatA directly binds to actin monomers, resulting in more extensive depolymerization of the actin network.45 Consequently, CytoB substantially reduced intracellular activity while producing only a modest reduction in intracellular mechanical resistance, whereas LatA strongly affected both the mechanical and active components of the fingerprint. This interpretation is supported by immunostaining experiments (Figure 2a and Appendix Figure 2), which revealed extensive reorganization of the actin cytoskeleton following both treatments. In both cases, filamentous actin structures became disrupted and the actin signal appeared redistributed into clustered structures. Compared with CytoB, LatA produced a more complete loss of recognizable filamentous actin, consistent with its stronger mechanical effect. As intracellular activity remained well above thermal levels after actin disruption, we next investigated the con-tribution of microtubule-associated processes. Since the probe particles are internalized by endocytosis and are typically enclosed within late endosomes or lysosomes, they are expected to interact strongly with microtubule-based transport processes. We therefore treated HeLa cells with 10 µg/mL Nocodazole to depolymerize the microtubule cytoskeleton. As expected, immunostainings for α-tubulin revealed a strong reduction of filamentous microtubule signal (Figure 2a, Appendix Figure 2). Nocodazole treatment caused a strong reduction of the solid-like prefactor A (Figure 2b) and overall reduction of storage and loss modulus (Figure 2d,g), and dramatically decreased intracellular activity from E = 23kBT to E = 4kBT (Figure 2h,j). Interestingly, this behavior differs from previous AFM studies probing cortical mechanics, which reported either no significant change in stiffness upon microtubule depolymerization46 or even a stiffening response.44 This discrepancy likely reflects the different mechanical compartments probed by the two methods. Whereas AFM predominantly probes the actin-rich cell cortex, our measurements characterize the mechanical environment ex-perienced by membrane-bound intracellular particles, which are expected to interact strongly with microtubule-based transport processes. Consistent with the pharmacological treatment, immunostaining confirmed the near-complete loss of the microtubule network following Nocodazole treatment, while revealing a qualitative increase in actin stress fibers (Figure 2a, Appendix Figure 2). Together, these findings support the interpretation that microtubule-associated processes make a major contribution to the active mechanical properties experienced by phagocytosed particles.

To this point, the data suggest that both filament types, actin and microtubules, contribute to the overall cytoplasmic mechanical properties, even though with different proportions. Yet, no treatment by itself was able to fully suppress intracellular activity. To test whether additional processes could contribute, we performed double disruption of actin and microtubules using the combined treatment with CytoB + Noc and LatA + Noc. Both conditions led to a reduction of cytoplasmic microtubules as well as dissolution of actin fibers and actin aggregation (Figure 2a and Appendix Figure 2). In both cases, the solid-like properties decreased beyond the levels previously observed by single disruptions (Figure 2b,d). The same can now also be observed for intracellular activity as quantified by E, which is reduced to about 2kBT for CytoB + Noc and down to almost thermal energy 1kBT for LatA + Noc treatment (Figure 2 h,i,j). The combined disruption of actin and microtubules is sufficient to suppress most of the intracellular activity experienced by phagocytosed particles.

These results demonstrate that the mechanical fingerprint can be used to dissect the effect of cytoskeletal drugs on different properties. At this point, it is important to stress that these experiments were performed on phagocytosed beads, likely engulfed in a membrane, which is an adequate proxy for membrane-bound organelles. The observed relations here can not be readily translated to inert particles. Other studies have found, for example, that the mechanical properties felt by inert probe particles are largely influenced by molecular crowding. Here, an increased intracellular activity led to a decreased overall stiffness.47, 48

Separating cell types by their mechanical fingerprint

The strong sensitivity of the solid-like prefactor A and the effective-energy prefactor E to cytoskeletal perturbations suggests that these parameters may be subject to cell-type-specific regulation. We therefore investigated whether cells with distinct biological functions and mechanical demands exhibit different mechanical fingerprints by comparing HeLa cells with two additional cell types (Figure 3a). First, we hypothesized that a muscle cell needs to have an increased intracellular mechanical stiffness for its function as a mechanical force generator. This suggests that muscle cells will be tuned to differ in the solid-like material properties with a higher prefactor A.

a) The mechanical fingerprint of 7 different cell types was determined by measuring their complex shear modulus and effective energy. b) Comparing HeLa cells to C2C12 muscle cells shows that muscle cells are overall stiffer (higher A). c) The comparison between HeLa cells and macrophages shows that macrophages have a higher intracellular activity (increased E) but are also more liquid-like and softer (increased α, β and decreased B) d-l). Using the fingerprint, active mechanical differences between cell types can be compared according to changes in the more solid-like material properties, the more liquid-like material properties, and intracellular activity.

As seen in Figure 3b, this is exactly what we found when comparing the cervical cancer HeLa cells with a murine-derived myoblast cell line (C2C12). The solid-like material stiffness, as quantified by A, significantly increased from about 13 to 20 (Figure 3b,d,e). Additionally, the fluid-like material prefactor B increased, indicating that the fluid-like material class also has more resistive properties (Figure 3b,g,h).

Next, we hypothesized that immune cells would have increased active forces in the cytoplasm and render it more liquid-like to be able to move more quickly through small constrictions as found in the extracellular space. This hy-pothesis was tested using an immune cell model. Macrophages were obtained from a hematopoietic progenitor cell line that expresses HoxB8 under an estrogen promoter.11, 13 Upon removal of estrogen, the cells differentiate within 3 days to macrophages that we term HoxB8 cells in the following for convenience. Again, our experimental analysis of the mechanical fingerprint is consistent with the prediction based on the biological function of macrophages (Figure 3c). Firstly, the effective energy prefactor E increased, indicating a raised intracellular activity. Secondly, the solid-like exponent α significantly increased, while the viscosity-defining prefactor B of the liquid-like material property collapsed (Figure 3f,h,k), indicating that the solid-like material class behaves more liquid-like, allowing the cytoplasm to quickly dissipate any large-scale deformation of the whole cell. Our results are in good agreement with measurements obtained previously on the whole cell level.49

These results suggest that differences in cellular function are associated with differences in mechanical fingerprint parameters. Muscle cells, requiring higher mechanical resistance, show a large solid-like prefactor, while immune cells that need to migrate quickly and deform easily are more liquid-like with lower viscous dissipation and generate a larger cytoplasmic active energy.

The mechanical fingerprint is different across cell types and species

Motivated by the finding that different cell types have distinct mechanical properties as quantified by the mechanical fingerprint, we wondered whether this approach could be used as a strategy to differentiate between cell types in terms of mechanical properties. In detail, we ask if at least one of the 6 parameters of the mechanical fingerprint is statistically different when comparing different cell types. Hence, we decided to determine the 6-parameter fingerprint for 7 different cell types (Figure 3a), of various functions and from various species. In addition to the already introduced human HeLa, murine C2C12 and HoxB8 cells, we measured the fingerprint of human lung epithelial cells (A549), a murine fibroblast-like colon carcinoma cell line (CT26), a canine kidney epithelial cell line (MDCK), and a non-invasive human breast cancer cell line (MCF7). Figure 3d-l summarizes the six fingerprint parameters across all seven investigated cell types. The observed trends are broadly consistent with previous studies of cell mechanics. For example, AFM measurements probing cortical stiffness reported a similar ordering of A549, MDCK, and MCF7 cells (A549: 1.6 kPa >MDCK: 0.56 kPa >MCF7: 0.25 kPa).50 As expected, the absolute stiffness of the cortex is much higher than that found in the cytoplasm. Notably, the same ordering is observed for the solid-like prefactor A (Figure 3e), despite the fact that our measurements probe the mechanical environment experienced by intracellular tracer particles rather than the cell cortex directly.

In a first analysis, we looked at all possible pair-wise combinations of cells to test our hypothesis that the 6 parameters are sufficient to differentiate between cell types. Indeed, in most of the 21 possible combinations we found at least one of the fingerprint parameters to be different (Figure 4a, Appendix 2.3). However, this analysis does not take into account the extent of the significant difference. To approach this question more systematically, we use a variant of the z-score, where we normalize the difference in the mean values by twice the geometric average of the standard deviation of the bootstrapped data. The z-score, hence, quantifies the extent of deviation between cells that is carried by a single parameter. For example, the z-score of A between HeLa and CT26 is: (5) zHeLa,CT26A=⟨AHeLa⟩−⟨ACT26⟩2σA,HeLa2+σA,CT262

a) Number of significantly different fingerprint parameters in pairwise cell type comparisons. b) Frequency of parameters showing the highest z-score. c) Distribution of z-score for all parameters for pairwise cell comparison. d) Correlation analysis between fingerprint parameters shows that not all parameters are varied independently. e) Explained variance ratio of the different principal components (PC) of the principal component analysis (PCA). f) Relative contribution of the fingerprint parameter to the first three principal components. PC1 mainly consists of α, β and log B, PC2 mainly consists of log E and γ and PC3 to 50% of log A. g) Plotting all cell types according to the first two principal components PC1 and PC2 shows that two parameters are already sufficient to distinguish between most cell types. h,i,j,k) Qualitative phase diagram of the active mechanical space. Activity, mainly captured by parameter E, resistance, dominated by A, and solid-liquid switching can be described as fluidity that is determined by principal component 1 (PC1) are varied among different cell types. Using this three-dimensional space allows identification of physical differences among different cell types.

Here, any value greater than 1 can be considered to mark a significant difference, and the larger the z-score, the better the two cells can be separated according to the respective parameter. We then calculated the z-score for all 6 fingerprint parameters across all 21 different pairs of cells. In a first step, we studied which of the parameters was best to differentiate between the cells. Figure 4b shows how often each parameter was the best discriminator. We see that E and A are the most important parameters to distinguish between different cell types as they together show the highest z-score in about 55% of the cases. As they mainly characterize the solid-like material properties and overall activity, this result implies that these properties are highly dependent on cell type and function. Additionally, we directly plot the z-scores of all cell pairs for the different parameters (Figure 4c). In contrast to the previous quantity, here the value of the z-scores becomes important. Despite E and A having most frequently the highest z-score, we also observed that the average z-score for all parameters was similar or smaller than 1, suggesting that the combination of all 6 parameters is relevant for characterizing the mechanical state of cells. The similarity in average z-scores may also indicate that there are correlations between these parameters, which could be due to underlying biological mechanisms.

To further explore this possibility, we looked at the correlation values (c.v.) (M 3.4) between the fingerprint parameters across all measurements as depicted in Figure 4d. We define an absolute c.v. below 0.3 as no (anti)correlation, between 0.3 and 0.6 as weak (anti)correlation, and above 0.6 as strong (anti)correlation. While for most parameter pairs no or only weak correlations are observed, we find a strong dependency between parameters B, β and α, as well as between E and γ. While α and β are positively correlated (0.85), both are negatively correlated with B (−0.71 and −0.79). A simultaneous variation of α, β, and B might be a way for the cell to switch between a solid- and fluid-like state. This finding also implies that, even though the two-power-law approach is able to perfectly describe the observed viscoelastic properties, a more simple model might also be able to fit the data, as some parameters are not fully independent. In this simpler model, the cytoplasm does might not consist of two independent classes of materials, but instead a deeper underlying mechanism might regulate the low-frequency elastic-like behavior just as the high-frequency more fluid-like behavior.

To see if we can indeed further reduce the number of parameters required to describe the active intracellular mechanical state, we performed a principal component analysis (PCA) with all three exponents α, β, and γ, and the logarithm of prefactors log A, log B and log E (M 3.5). The explained variance ratio of each component is presented in Figure 4e. Already 93% of the variance can be explained using only the first 3 components, and 78.4% can even be explained with just 2 components. This supports the hypothesis that few underlying biological mechanisms act as “hyperparameter”, which tune the mechanical fingerprint of the intracellular space.

Each principal component is a superposition of the normalized fingerprint parameters. We therefore wondered which fingerprint parameters dominate the first 3 principal components. Looking at the relative contributions (Figure 4f, Appendix 2.1), we find that the first component PC1, explaining 51.5% of all variance in data, mainly consists of fingerprint parameters α, B, and β (22%, 24%, 24%). Finding these parameters combined in a single principal component is expected, as they also show strong correlation (Figure 4d), but strikingly, even though neither parameter by itself appeared to be dominant in characterizing the intracellular state in the previous analysis (Figure 4b,c), the combined component PC1 seems to be the most relevant principal component. The second component PC2 consists mainly of the fingerprint parameters related to the intracellular activity E and γ (30.3% and 32%), and the third one largely consists of parameter A (50%), characterizing the overall resistance of the cytoplasm felt by phagocytosed particles.

To determine whether PC1 and PC2 are already sufficient to distinguish between cell types, we plot the cell types according to these two parameters (Figure 4g). We observe that most cell types appear to separate from one another, indicating that PC1 and PC2 are effective in distinguishing between them.

As the first three principal components are sufficient to distinguish most of the investigated cell types, we sought to assign a physical interpretation to them. The first principal component, PC1, is dominated by the parameters α, β, and B. Because these parameters are strongly correlated across the investigated cell types, an increase in PC1 corresponds to a transition from a more elastic, solid-like response toward a more dissipative and fluid-like mechanical state. We therefore interpret PC1 as a measure of cellular fluidity. In contrast, the second principal component is dominated by the intracellular activity parameter E and γ. As the overall amplitude of effective energy is mainly influenced by a variation in E and less by γ, an increase in PC2 reflects an increase in intracellular activity. Lastly, as the third component is mainly represented by parameter A, which in turn is typically varied over an order of magnitude higher values than B, a change of this component can be attributed to an overall change of absolute complex modulus and thus a change of overall resistance.

To avoid this ambiguity of having principal components consisting of multiple parameters while preserving the main features revealed by the principal component analysis, we propose a qualitative three-dimensional phase space spanned by intracellular activity (primarily represented by log E), cellular resistance (primarily represented by log A), and fluidity (represented by PC1) (Figure 4h).

Mapping the different cell types into this phase space reveals distinct mechanical phenotypes (Figure 4i,j,k). HoxB8 macrophages exhibit the highest intracellular activity while simultaneously occupying the most fluid-like region of the phase space and, together with MCF7 cells, displaying comparatively low resistance. In contrast, C2C12 myoblasts are characterized by high resistance and one of the most solid-like mechanical states. These observations suggest that different cell types occupy distinct regions of the active-mechanical phase space, potentially associated with their physiological functions.

Having established that different cell types can be distinguished by their active mechanical state, an important next question is how these mechanical properties relate to cellular function. In particular, it would be interesting to determine whether parameters such as migration speed, contractility, or proliferation correlate with specific regions of the phase space. However, quantitative functional data are highly dependent on experimental conditions, including culture medium, substrate properties, and assay design, making direct comparisons across the investigated cell types difficult. As one example, wound-healing experiments have reported higher migration speeds for HeLa cells than for MCF7 cells,51 which is qualitatively consistent with the higher intracellular activity observed for HeLa cells in our measurements. To obtain a first indication of possible structure-function relationships, we grouped the investigated cell types according to species, epithelial or mesenchymal character, cancer status, metastatic potential, and presumed migratory potential (Appendix Figure 3 Appendix Table 17). However, no clear correlations emerged from this analysis, emphasizing the need for future studies that combine intracellular rheological measurements with quantitative functional assays.

Conclusion

In this study, we introduced a mechanical fingerprint that quantitatively describes the active mechanical state of the cy-toplasm using only six parameters. By combining active and passive intracellular microrheology with a minimal physical description, we reduced the complex frequency-dependent viscoelastic and active properties of living cells to a compact parameter set that captures intracellular mechanics across a broad range of timescales.

Using pharmacological perturbations of the cytoskeleton, we demonstrated that individual fingerprint parameters reflect distinct physical aspects of the intracellular environment. While disruption of actin filaments, especially using Latrunculin A, primarily affected the solid-like mechanical response, microtubule depolymerization strongly reduced intracellular activity. Combined disruption of both filament systems reduced active fluctuations to nearly thermal levels, illustrating that the fingerprint can disentangle different contributions to intracellular active mechanics. Comparison with previous AFM studies revealed that actin depolymerization induces qualitatively similar softening responses at both the cortical and intracellular levels. In contrast, microtubule depolymerization produced markedly different effects, highlighting that intracellular microrheology using phagocytosed beads probes a mechanical compartment distinct from the actin-rich cell cortex and is particularly sensitive to microtubule-associated intracellular processes.

Extending the analysis across multiple cell types revealed that the mechanical fingerprint is not randomly distributed but instead exhibits characteristic differences between cell types. Muscle cells exhibited increased mechanical resistance and more solid-like properties, whereas macrophages displayed enhanced activity and increased fluidity. More generally, the statistical analysis revealed that the fingerprint parameters do not vary independently across the investigated cell types. Instead, changes in individual parameters are accompanied by systematic changes in others, suggesting that intracellular mechanical states occupy a constrained region of parameter space.

This coordinated tuning is reflected in a low-dimensional active mechanical phase space. Principal component analysis identified fluidity as a dominant mode of variation, while activity and mechanical resistance emerged as additional key descriptors. Together, these quantities define a three-dimensional framework that allows different cell types to be positioned according to their intracellular mechanical state. We propose that this phase space provides a useful basis for relating intracellular mechanics to biological function.

Several limitations should be considered when interpreting the presented results. The measurements were performed on phagocytosed probe particles that are likely enclosed within membrane-bound compartments and therefore probe the mechanical environment experienced by endosomes or lysosomes rather than the bulk cytoplasm directly. Consequently, the reported fingerprint reflects the active mechanics of membrane-associated intracellular structures and may differ from measurements obtained using inert tracer particles. Here, the mechanical properties are to a large extent influenced by the actin and microtubule cytoskeleton, while microtubule-associated processes emerge as major contributors to the measured fingerprint. Besides cytoskeletal organization, metabolic state, molecular crowding, and intracellular transport dynamics are likely to contribute to the observed fingerprint. Other studies probing inert particles in the cytoplasm have shown that intracellular mechanics can be strongly affected by molecular crowding.47, 48 Future studies combining intracellular rheology with metabolic perturbations or direct measurements of cellular energy metabolism will be required to disentangle the relative contributions of these factors.

Taken together, our results establish the mechanical fingerprint as a quantitative framework for characterizing intra-cellular active mechanics. We anticipate that this approach will enable systematic comparisons between cell types and physiological states and provide new opportunities to investigate how intracellular mechanics relate to cellular function and disease.

Material and Methods

Active and passive microrheology in cells

In order to conduct active and passive microrheology measurements, we utilized the previously described optical tweezers setup (see Figure 1a). Cells were prepared following the protocols outlined in Materials and Methods M 4. Throughout the experiments, the cells were maintained at 37 °C and 5% CO2 to ensure physiological conditions. We investigated a minimum of N=58 cells for each cell type or condition (see Appendix 2.2) over the course of at least n=3 consecutive days. For each cell, one particle was investigated. Data points were recorded at a sampling rate of 65 536 Hz.

Active microrheology

In active microrheology, the 808 nm trapping laser is employed to oscillate with a variable driving frequency fD, exerting a sinusoidal force F (t) onto a probe particle located inside a cell. Simultaneously, a second position detection laser records the position x(t) of the particle. The relationship between the particle’s position and the applied force is determined by the response function χ(t), as given by: (6) x(t)=∫−∞tχ(t−t′)F(t′)dt′.

In the frequency domain, this convolution can be evaluated as a product, providing access to the response function: (7) χ(f)=x~(f)F~(f), which characterizes the viscoelastic properties of the probed region. Here, x~ and F~ denote the Fourier transform of position and force.

To obtain the frequency-dependent response function, the probe particle is successively driven at different driving frequencies (fD ∈ 1Hz, 2Hz, 4Hz, 8Hz, 16Hz, 32Hz, 64Hz, 128Hz, 256Hz, 512Hz, 1024Hz). At each frequency, a minimum of 3 periods was recorded. For frequencies above 2Hz, we recorded a maximum duration of 1 second. Utilizing the generalized Stokes-Einstein relation:52 (8) G∗(f)=16πRχ(f), where R represents the particle radius, we determine the complex shear modulus G*(f).

Passive microrheology

In passive micro rheology measurements, free particle fluctuations inside the cells are recorded without application of external forces. In this study, we employ the weak detection laser and back focal plane interferometry to track the bead displacement with high spatial and temporal precision. The particle position x(t) is tracked 3 consecutive times over a time span of 10 s for each cell investigated. For each time series x(t), we calculate the power spectral density (PSD) (9) C~(f)=|x~(f)|2 and average it over the 3 repetitions.

In equilibrium systems, the PSD is linked to the mechanical properties of the system via a generalized fluctuation-dissipation theorem (FDT):52, 53 (10) C~(f)=kBTχ′′(f)πf

Here kBT is the thermal energy of the system and χ″(f) the imaginary part of the response function. In passive systems, this equation is valid and C(f) can be used to determine the mechanical properties of the system. Living cells, however, are non-equilibrium systems. Metabolic energy leads to stronger particle fluctuations C(f) than one would expect for a passive systems. We define an effective energy by adding an active cellular energy Eactive to the thermal energy: (11) EEff(f)=Eactive(f)+kBT

This is the energy required to explain the particle fluctuations C(f) given the mechanical properties χ″(f) measured in active microrheology. This yields a new adapted version of the FDT: (12) C~(f)=EEff(f)χ′′(f)πf

By separately measuring χ″(f) in active microrheology and C(f) in passive microrhelogy, we gain access to the effective energy of the system and thereby the level of non-equilibrium: (13) EEff(f)=C(f)χ′′(f)πfkBT

The effective energy quantifies the level of activity in the system. Any deviation from 1 is indicative for non-equilibrium.

Optical setup

In this study, we use an optical tweezers setup that has been previously described34 to perform active and passive microrheology measurements and apply well-defined forces onto phagocytozed probe particles in living cells. Briefly, the setup is based on a home-built bright-field microscope equipped with a 60x NA 1.2 Objective, where we employ two different lasers. A 808 nm laser (LU0808M250, 808 nm, 250 mW Lumics GmbH, Berlin,331 Germany) is operated at moderate power (70 mW at sample plane) to allow for stable optical trapping and force application to probe particles at the focal plane. The lateral position of this laser can be precisely controlled with a piezo-based tilting mirror. Using a high NA=1.4 condenser, most of the transmitted laser light is collected. By projecting the back focal plane of the condenser onto a position sensitive detector (PSD), the trapping forces can be directly quantified.54 A second, stationary, laser (L976-PAG500, 976 nm, 500 mW, Thorlabs, New Jersey, USA) is operated at low laser power (0.3 mW at sample plane) to not apply any relevant trapping forces. This laser is utilized to monitor the bead displacement via back focal plane interferometry55 with nm precision. A schematic of this setup is depicted in Figure 1a.

Statistical analysis

Bootstrapping

Bootstrapping is employed to get an estimate for the distribution of the mean cellular active mechanical properties. The following scheme is used:

We have i experimental conditions, either different cell-types or drug treatments. For each condition i, we inves-tigate Ni different cells. For each cell, we obtain curves for G′,G″ and EEff

For each condition i, the bootstrapping data set is generated. Ni samples are drawn with replacement with consecutive mean calculations. This process is repeated Nbootstrap=10.000 times for each quantity of interest.

The resulting distribution of curves captures the distribution of the means.

Each curve is subsequently fit with the respective model, which in turn also yields the distribution of the mean fit parameters.

Significance tests

Statistical comparisons between fit parameters obtained from the bootstrapping scheme (Section M 3.1) were performed as follows;56 a schematic is provided in Figure 1.

For each pairwise comparison of two groups G1 and G2, we exploit that bootstrapping yields a full sampling distri-bution of each fit parameter rather than a single point estimate. Let {θ1(b)} and {θ2(b)} denote the bootstrap distributions of a given parameter for the two groups. We form the difference distribution (14) Δ(b)=θ2(b)−θ1(b), and estimate the two-sided p-value as (15) p=2min(#{Δ(b)<0}B,1−#{Δ(b)<0}B), where B is the number of bootstrap replicates. Intuitively, this tests whether zero is contained in the bulk of the difference distribution. If the two groups differ substantially, the entire distribution is shifted away from zero, and the p-value is small.

Because multiple pairwise comparisons are performed simultaneously for each parameter, we control the familywise error rate using the Holm–Bonferroni sequential correction.57 Significance levels are then assigned to the adjusted p-values as: p ≤ 0.001 (*∗*), p ≤ 0.01 (**), p ≤ 0.05 (*), and p > 0.05 (n.s.).

Z-score

We use an adapted z-score to quantify how different a certain parameter is between two different cell types. (16) z=⟨AHeLa⟩−⟨ACT26⟩2σA,HeLa2+σA,CT262

Here σ is the standard deviation SD.

Correlation analysis

Correlation analysis was performed using the corrcoef function of the python package numpy.58

Principal component analysis PCA

In order to see if the complexity of the 6-parameter fingerprint log E, γ, log A, α, log B, β can be further reduced, we per-formed a principal component analysis on the data. To this end, we utilized the function sklearn.decomposition.PCA from the Python module scikit-learn.59 The data was normalized prior to the PCA using the sklearn.preprocessing.StandardSca method.

Cell culture and bead insertion

A549, C2C12, CT26, HeLa, and MDCK cells were cultured in Dulbecco’s modified Eagle medium (DMEM, Capricorn) supplemented with 1% Penicillin Streptomycin (Gibco) and 10% fetal bovine serum (FBS, Sigma-Aldrich) at a temperature of 37°C and 5% CO2.

HoxB8 cells, provided by the Institute for Immunology at the University of Münster, Germany, were cultured according to a protocol described elsewhere.11 To induce differentiation into macrophages, HoxB8 cells were suspended in differentiation medium three days prior to the experiment. After this, the same procedure as for the other cell lines, with the exception of using EDTA instead of Trypsin, was applied.

Cells were split when they reached near full confluence and seeded onto Fibronectin-coated glass cover slips (22×55×0.15 mm VWR). Medium was then aspirated, and a fresh 1:10,000 dilution of 1 µm carboxylated beads (FCDG006, Bangslabs) in medium was added to the sample. Cells were incubated for up to 15 hours to allow for adequate bead phagocytosis. Then, cells were washed with PBS to remove any residual extracellular beads. Following aspiration of PBS, a cover slip was fixed to the estimated cell spreading area using two layers of 200 µm adhesive tape (DST1950, Thorlabs, New Jersey, USA).

The sample chamber was filled with CO2-independent medium (CO2 Independent Medium, Gibco™) to ensure optimal cell viability during the experiments.

Pharmacological treatment

Cytochalasin B (Sigma-Aldrich), Latrunculin A (Sigma-Aldrich), and Nocodazole (Sigma-Aldrich) were used to perturb the cytoskeleton. Cytochalasin B inhibits actin polymerization by capping filament ends, whereas Latrunculin A binds to actin monomers and thereby induces actin depolymerization. Nocodazole disrupts the microtubule network by binding to β-tubulin and inhibiting microtubule polymerization.

Cells were subjected to one of five pharmacological treatments: Cytochalasin B (CytoB 10 µg mL−1), Latrunculin A (LatA 1µM), Nocodazole (Noc 10 µg mL−1), Cytochalasin B + Nocodazole (CytoB+Noc), or Latrunculin A + Noco-dazole (LatA+Noc). All treatments were applied for 60 min prior to the measurements. Combined treatments were performed using the same drug concentrations as in the respective single-drug conditions.

Immunofluorescence staining and imaging

To validate the cytoskeletal perturbations, cells were fixed and subjected to immunofluorescence staining for F-actin, microtubules, and nuclei. Following drug treatment, cells were washed three times with PBS and fixed with 4% paraformaldehyde (PFA) for 15 min at room temperature. Samples were subsequently washed with PBS and per-meabilized in PBS containing 10% goat serum and 0.2% Triton X-100 for 10 min. Cells were then incubated in a blocking solution consisting of 10% goat serum and 0.2% Tween-20 in PBS.

Microtubules were stained using a mouse anti-α-tubulin primary antibody (A11126, Invitrogen; 1:100 dilution) for 2 h at room temperature, followed by incubation with an Alexa Fluor 568-conjugated goat anti-mouse secondary antibody (A21124, Invitrogen; 1:500 dilution) for 90 min. F-actin was labeled using Phalloidin-iFluor 647 (ab176759, Abcam; 1:500 dilution), and nuclei were stained with Hoechst (H1399, Thermo Fisher Scientific; 1:2500 dilution). All staining steps were performed protected from light. After staining, samples were washed three times with PBS and stored at 4°C until imaging.

Fluorescence image stacks were acquired using a spinning-disk confocal microscope (CSU-W1, Yokogawa) mounted on a Nikon Ti-1 microscope equipped with a 60× WI objective. Z-stacks were recorded with a spacing of 270 nm between adjacent planes.